A primer to numerical simulations: The perihelion motion of Mercury

### 2017-05-23 tex/MercuryNumerical.tex
Added some links for ball in the box (Christian)

In python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
Next, we set up the initial vectors which will describe the motion in the two-dimensional plane.  We will build on the existing \python module: \texttt{Vpython}, which provides implementation for the vectorial implementation and animation.  The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions, the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rS0 = vector(0, 0, 0)
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.  In first approximation we can numerically express the updates of the position and the velocity by
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

@@

\bibliographystyle{plain}
\bibliography{MercuryNumerical}

>>>>

In python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
Next, we set up the initial vectors which will describe the motion in the two-dimensional plane.  We will build on the existing \python module: \texttt{Vpython}, which provides implementation for the vectorial implementation and animation.  The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions, the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.  In first approximation we can numerically express the updates of the position and the velocity by
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

@@

\bibliographystyle{alpha din}
\bibliography{MercuryNumerical}
